---
title: "Final Document"
author: "Yifan Zhang, Xiang Li, Shuting Li, Ranfei Xu"
date: "2021/11/22"
output:
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(drat)
library(hurricaneexposuredata)
library(hurricaneexposure)
library(weathermetrics)
library(geoR)
library(dplyr)
library(gstat)
library(gridExtra)
library(dplyr)
library(magrittr)
library(maps)
library(geosphere)
library(sp)
Sys.setenv(LANGUAGE = "en")
```

# Introduction

__Hurricane Humberto__ is a tropical storm on September 12, 2007. Humberto is the first hurricane to make landfall in the United States since Wilma struck South Florida in October 2005 and the first hurricane to make landfall along the Texas coast since Rita struck the Texas-Louisiana area in September 2005. In this report, we are going to map Humberto's weather effect and compare average wind speed by using buoy’s data from National Data Buoy Center.

## Plotting Storm Tracks and Selected Buoys
```{r}
# data preparation
addRepo("geanders")
hurr_tracks <- hurricaneexposuredata::hurr_tracks
hurr_Humberto2007         <- hurricaneexposuredata::hurr_tracks %>% subset(storm_id == "Humberto-2007")
rain_Humberto2007         <- hurricaneexposuredata::rain %>% subset(storm_id == "Humberto-2007")
wind_Humberto2007         <- hurricaneexposuredata::storm_winds %>% subset(storm_id == "Humberto-2007")
closest_dist_Humberto2007 <- hurricaneexposuredata::closest_dist %>% subset(storm_id == "Humberto-2007")

# 42019 Location: 27.910N 95.345W
# 42035 Location: 29.232N 94.413W
# SRST2 Location: 29.683N 94.033W

y <- c(27.910, 29.232, 29.683) 
x <- c(-95.345, -94.413, -94.033)
bouys_location <- data.frame(x,y)

```

```{r}
# the way below is wrong
# track <- map_tracks(storms = "Humberto-2007")
# track + geom_point(data = bouys_location, aes(x=x,y=y), size = 2, shape = 23,  fill = "yellow")

region <- c("Arkansas","Louisiana", "Mississippi", "Texas")  
# you can also add the whole state 
# region <- c("alabama", "arkansas", "connecticut", "delaware", "district of columbia",
#                 "florida","georgia", "illinois", "indiana","iowa", 
#                 "kansas", "kentucky", "louisiana","maine", "maryland",
#                 "massachusetts", "michigan", "mississippi", "missouri","new hampshire", 
#                 "new jersey", "new york", "north carolina", "ohio", "oklahoma", 
#                 "pennsylvania", "rhode island", "south carolina","tennessee", "texas",
#                 "vermont","virginia", "west virginia", "wisconsin")  
state    <- map_data(map = "state", region = region)
state_text <- state %>%
  group_by(region) %>% 
  summarize(long = mean(long), lat = mean(lat), .groups = "keep")
# plot track and buoy
track <-
  ggplot() + 
  geom_polygon(data = state, aes(x = long, y = lat, group = group),
               color="black", fill="white", size = 0.2, alpha = 0.3) +
  geom_path(aes(x = hurr_Humberto2007$longitude, y = hurr_Humberto2007$latitude), color = "red4") +
  geom_point(data = bouys_location, aes(x=x,y=y), size = 2, shape = 23,  fill = "yellow")+
  geom_text(aes(x = long, y = lat, label = region), data = state_text, size = 6)+
  ggtitle("Humberto-2007") +
  theme(plot.title=element_text(hjust=0.5),
        panel.background=element_blank(),
        panel.border=element_blank(),
        axis.title=element_blank(),
        axis.text=element_blank(),
        axis.ticks=element_blank())
track
```

\newpage
# Maps for Hurricane Humberto from “Hurricane Exposure” Package

We generated maps of __Hurricane Humberto-2007__ from the “Hurricane Exposure” package, which included rain maps, wind speed, and duration maps for sustained wind and wind gusts.

Here is the difference between sustained wind and wind gusts

|Name          | Definition|
| :--:         | :----- |
|sust          |A sustained wind is defined as the average wind speed over two minutes. |
|gust          |A sudden burst in wind speed is called the wind gusts and typically lasts under 20 seconds. |

On average, the diameter of hurricane-force winds is about 100 miles (160 kilometers). Slower speed winds further out from the storm center increase the storm system's diameter on average 300 to 400 miles across. And the eye of the hurricane is 30 miles in diameter. Therefore, we only choose exposure counties, which are 117 counties in total.


```{r,warning=FALSE,message=FALSE}
# function
hmapper <- function(hurr){
    
    rmap = map_counties(storm = hurr, metric = "rainfall") +
            ggtitle("Rainfall") +
            theme(plot.title = element_text(hjust = 0.5))

    wmap = map_counties(storm = hurr, metric = "wind") +
            ggtitle(hurr) +
            theme(plot.title = element_text(hjust = 0.5))

    expos = map_rain_exposure(storm =hurr, 
                  rain_limit = 100, 
                  dist_limit = 500, 
                  days_included =-5:3) +
                    ggtitle(hurr) +
                    theme(plot.title = element_text(hjust = 0.5))

    ml <-  list(rmap, wmap, expos)
    names(ml) <- c("rmap", "wmap", "expos")
    
    return(ml)
}


county_center <- hurricaneexposuredata::county_centers

# plot
plot_Humberto2007 <- hmapper("Humberto-2007")
plot_Humberto2007$rmap

hurr = "Humberto-2007"
# set dist_limit = 160 to define exposure.
map_rain_exposure(storm = hurr,
      rain_limit = 0,
      dist_limit = 160,
      days_included =-5:3) +
        ggtitle("Counties Exposed by Rain") +
        theme(plot.title = element_text(hjust = 0.5))
# distance visualization
map_counties(storm = hurr,
             metric = "distance",
             add_track = TRUE)+
             ggtitle("Distance")+
             theme(plot.title = element_text(hjust = 0.5))

m1 <- map_counties(storm = hurr, 
             metric = "wind",
             wind_var = "vmax_sust", 
             add_track = TRUE)+
  ggtitle("Wind vmax_sust")
m2 <- map_counties(storm = hurr, 
             metric = "wind",
             wind_var = "vmax_gust", 
             add_track = TRUE) +
   ggtitle("Wind vmax_gust")
m3 <- map_counties(storm = hurr, 
             metric = "wind",
             wind_var = "sust_dur", 
             add_track = TRUE)+
   ggtitle("Wind sust_dur")
m4 <- map_counties(storm = hurr, 
             metric = "wind",
             wind_var = "gust_dur", 
             add_track = TRUE)+
     ggtitle("Wind gust_dur")
grid.arrange(m1,m2,ncol=2)
grid.arrange(m3,m4,ncol=2)
```


\newpage
# Spatial Analysis

Here are the variogram analysis generated from the “Hurricane Exposure” Package

+ Rain

```{r}
# subset rain and wind data to these 117 counties
expo_county <- closest_dist_Humberto2007 %>% subset(storm_dist < 160) # filter 117 exposure counties 
expo_rain   <- rain_Humberto2007 %>% inner_join(expo_county, by = c("fips", "storm_id", "usa_atcf_id")) %>% 
    select(fips, storm_id, lag, precip, precip_max)
expo_wind   <- wind_Humberto2007 %>% inner_join(expo_county, by = c("fips", "storm_id", "usa_atcf_id")) %>% 
    select(fips, storm_id, vmax_gust, vmax_sust, gust_dur, sust_dur)
# 1053/9 # exactly 117 counties

# prepare for variogram analysis(add county_name, latitude and longitude)
expo_rain <- expo_rain %>% inner_join(county_center, by = "fips") %>% 
  dplyr::select(fips, storm_id, county_name, state_name, latitude, longitude, lag, precip, precip_max)
expo_wind <- expo_wind %>% inner_join(county_center, by = "fips") %>% 
  dplyr::select(fips, storm_id, county_name, state_name, latitude, longitude, vmax_gust, vmax_sust, gust_dur, sust_dur)

## variogram analysis
# use Haversine formula to calculate distance
rain_county <- expo_rain %>% 
  select(fips, storm_id, county_name, state_name, latitude, longitude) %>% unique()
expo_rain_mean <- expo_rain %>% group_by(fips) %>% 
  summarise(precip = mean(precip), precip_max = mean(precip_max)) %>% 
  inner_join(rain_county, by = "fips")
coordinates(expo_rain_mean) <- ~ latitude + longitude

# variogram of precip
variog_rain_precip <- variogram(log(precip+1) ~ 1, data = expo_rain_mean, width = .03)
fit_rain_precip    <- fit.variogram(variog_rain_precip, vgm(c(c("Exp", "Mat", "Sph"))))
line_rain_precip <- variogramLine(vgm(psill = fit_rain_precip$psill[2], "Sph", 
                       range = fit_rain_precip$range[2], nugget = 0), 3)
v1 <- ggplot(variog_rain_precip) + 
  geom_point(aes(x = dist, y = gamma, color = dist)) + 
  geom_line(data = line_rain_precip, aes(x = dist, y = gamma), color = "red") + 
  ggtitle("Variogram -- Precipitation") + theme_bw() + 
  theme(legend.position = "none")

# variogram of precip_max
variog_rain_precip_max <- variogram(log(precip_max+1) ~ 1, data = expo_rain_mean, width = .03)
fit_rain_precip_max    <- fit.variogram(variog_rain_precip_max, vgm(c(c("Exp", "Mat", "Sph"))))
line_rain_precip_max <- variogramLine(vgm(psill = fit_rain_precip_max$psill[2], "Sph", 
                       range = fit_rain_precip_max$range[2], nugget = 0), 3)
v2 <- ggplot(variog_rain_precip_max) + 
  geom_point(aes(x = dist, y = gamma, color = dist)) + 
  geom_line(data = line_rain_precip_max, aes(x = dist, y = gamma), color = "red") + 
  ggtitle("Variogram -- Max Precipitation") + theme_bw() + 
  theme(legend.position = "none")
grid.arrange(v1,v2,ncol=2)
```


Variogram interpretation:

Among 117 counties, the correlation between rain __Precipitation__ and the correlation between __Max-Precipitation__ are increasing as the distance between two random counties is increasing.

+ Wind

```{r}
expo_wind_mean <- expo_wind
coordinates(expo_wind_mean) <- ~ latitude + longitude

# variogram: gust wind
variog_wind_gust <- variogram(log(vmax_gust+1) ~ 1, data = expo_wind_mean, width = .03)
fit_wind_gust    <- fit.variogram(variog_wind_gust, vgm(c(c("Gau"))))
line_wind_gust <- variogramLine(vgm(psill = fit_wind_gust$psill[2], "Gau", 
                                    range = fit_wind_gust$range[2], 
                                    nugget = fit_wind_gust$psill[1]), 3)
plot_wind_gust <- ggplot(variog_wind_gust) + 
  geom_point(aes(x = dist, y = gamma, color = dist)) + 
  geom_line(data = line_wind_gust, aes(x = dist, y = gamma), color = "red") + 
  ggtitle("Variogram -- Gust wind") + theme_bw() + 
  theme(legend.position = "none")

# variogram: sustained wind
variog_wind_sust <- variogram(log(vmax_sust+1) ~ 1, data = expo_wind_mean, width = .03)
fit_wind_sust    <- fit.variogram(variog_wind_sust, vgm(c(c("Gau"))))
line_wind_sust <- variogramLine(vgm(psill = fit_wind_sust$psill[2], "Gau", 
                                    range = fit_wind_sust$range[2], 
                                    nugget = fit_wind_sust$psill[1]), 3)
plot_wind_sust <- ggplot(variog_wind_sust) + 
  geom_point(aes(x = dist, y = gamma, color = dist)) + 
  geom_line(data = line_wind_sust, aes(x = dist, y = gamma), color = "red") + 
  ggtitle("Variogram -- sustained wind") + theme_bw() + 
  theme(legend.position = "none")

# variogram: gust wind duration
variog_wind_gust_dur <- variogram(log(gust_dur+1) ~ 1, data = expo_wind_mean, width = .03)
fit_wind_gust_dur    <- fit.variogram(variog_wind_gust_dur, vgm(c(c("Gau"))))
line_wind_gust_dur <- variogramLine(vgm(psill = fit_wind_gust_dur$psill[2], "Gau", 
                                    range = fit_wind_gust_dur$range[2], 
                                    nugget = fit_wind_gust_dur$psill[1]), 3)
plot_gust_dur <- ggplot(variog_wind_gust_dur) + 
  geom_point(aes(x = dist, y = gamma, color = dist)) + 
  geom_line(data = line_wind_gust_dur, aes(x = dist, y = gamma), color = "red") + 
  ggtitle("Variogram -- gust wind duration") + theme_bw() + 
  theme(legend.position = "none")

# variogram: sustained wind duration
variog_wind_sust_dur <- variogram(log(sust_dur+1) ~ 1, data = expo_wind_mean, width = .03)
fit_wind_sust_dur    <- fit.variogram(variog_wind_sust_dur, vgm(c(c("Gau"))))
line_wind_sust_dur <- variogramLine(vgm(psill = fit_wind_sust_dur$psill[2], "Gau", 
                                        range = fit_wind_sust_dur$range[2], 
                                        nugget = fit_wind_sust_dur$psill[1]), 3)
plot_sust_dur <- ggplot(variog_wind_sust_dur) + 
  geom_point(aes(x = dist, y = gamma, color = dist)) + 
  geom_line(data = line_wind_sust_dur, aes(x = dist, y = gamma), color = "red") + 
  ggtitle("Variogram -- sustained wind duration") + theme_bw() + 
  theme(legend.position = "none")

grid.arrange(plot_wind_gust, plot_wind_sust, plot_gust_dur, plot_sust_dur,ncol=2,nrow=2)

```


Among 117 counties, the correlation between wind_gust, the correlation between wind_sust, the correlation between gust_dur, and the correlation between sust_dur are increasing as the distance between two random counties is increasing.

The reason why we do not have a sill in these variogram analyses is that we only choose 117 counties that are exposed by Hurricane Humberto.

#  Exploration Data Analysis from Buoys 

We generated plots from buoys 42019, 42035, and SRST2, which were significantly influenced by Humberto.

Here is the date and time when Hurricane Humberto was closest to each buoy. 

```{r,warning=FALSE,message=FALSE}
# insert buoys data
library(tidyverse)
library(magrittr)
library(lubridate)


#Here is the Measurement description: https://www.ndbc.noaa.gov/measdes.shtml#stdmet


#Only 42019, 42035 and SRST2 were significantly influenced by Humberto.44013

# These three datasets are Standard Meteorological Data
#https://www.ndbc.noaa.gov/download_data.php?filename=42019h2007.txt.gz&dir=data/historical/stdmet/
#https://www.ndbc.noaa.gov/download_data.php?filename=42035h2007.txt.gz&dir=data/historical/stdmet/
#https://www.ndbc.noaa.gov/download_data.php?filename=srst2h2007.txt.gz&dir=data/historical/stdmet/

buoys_id <- c("42019", "42035", "SRST2")
buoys_id %<>% tolower() 
url1 <-  "https://www.ndbc.noaa.gov/view_text_file.php?filename="
url2 <- ".txt.gz&dir=data/historical/stdmet/"
urls <- str_c(url1,buoys_id, "h2007", url2, sep = "")
filenames <- str_c("dt_", buoys_id, sep = "")

N <- length(urls)

for (i in 1:N){
  suppressMessages(  ###  This stops the annoying messages on your screen.
   file <- read_table(urls[i], col_names = TRUE)
  )
   file$date_time <- make_datetime(year = 2007, month = as.integer(file$MM), day = as.integer(file$DD), hour = as.integer(file$hh), min = as.integer(file$mm))
   file <- file[file$date_time>"2007-09-06 0:0:00 EST",]
   file <- file[file$date_time<"2007-09-18 23:0:00 EST",]
   file <- file[!is.na(file$`#YY`),]
  assign(filenames[i], file)
}

# delete the "mm" column
for (i in 1:N){
  file <- get(filenames[i])
  assign(filenames[i],file[,c(1:4,6:19)])
}
# summarise daily data(dt_42019)
raw_42019   <- dt_42019 %>% select(date_time, WSPD) %>% 
  add_column(buoy = rep("buoy_42019", length(dt_42019$date_time)), .after = 0)
colnames(raw_42019)[3] <- c("average_wind_speed")
raw_42019$average_wind_speed <- as.numeric(raw_42019$average_wind_speed)

# summarise daily data(buoy_42035)
raw_42035   <- dt_42035 %>% select(date_time, WSPD) %>% 
  add_column(buoy = rep("buoy_42035", length(dt_42035$date_time)), .after = 0)
colnames(raw_42035)[3] <- c("average_wind_speed")
raw_42035$average_wind_speed <- as.numeric(raw_42035$average_wind_speed)

# summarise daily data(buoy_srst2)
raw_srst2   <- dt_srst2 %>% select(date_time, WSPD) %>% 
  add_column(buoy = rep("buoy_srst2", length(dt_srst2$date_time)), .after = 0)
colnames(raw_srst2)[3] <- c("average_wind_speed")
raw_srst2$average_wind_speed <- as.numeric(raw_srst2$average_wind_speed )

# put them into one Dataframee
buoy_raw <- rbind(raw_42019, raw_42035, raw_srst2)
buoy_location <- data.frame(buoy = unique(buoy_raw$buoy), 
                            longitude  = bouys_location$x, 
                            latitude   = bouys_location$y)
buoy_raw <- buoy_raw %>% inner_join(buoy_location, by = "buoy")

# calculate Haversine distance
hurr_Humberto2007 %<>% mutate(dist_42019 = NA, dist_42035 = NA, dist_srst2 = NA)
for(i in 1:  dim(hurr_Humberto2007)[1]){
  hurr_Humberto2007$dist_42019[i] <- distHaversine(c(hurr_Humberto2007$longitude[i], hurr_Humberto2007$latitude[i]), 
                                                   c(buoy_location$longitude[1], buoy_location$latitude[1]), r=6378137)
  hurr_Humberto2007$dist_42035[i] <- distHaversine(c(hurr_Humberto2007$longitude[i], hurr_Humberto2007$latitude[i]),
                                                   c(buoy_location$longitude[2], buoy_location$latitude[2]), r=6378137)
  hurr_Humberto2007$dist_srst2[i] <- distHaversine(c(hurr_Humberto2007$longitude[i], hurr_Humberto2007$latitude[i]),
                                                   c(buoy_location$longitude[3], buoy_location$latitude[3]), r=6378137)
}
buoy_location %<>% mutate(closest_date = rep(NA, dim(buoy_location)[1]))

# set the closest_date
buoy_location$closest_date[1] <- hurr_Humberto2007$date[which(hurr_Humberto2007$dist_42019 == min(hurr_Humberto2007$dist_42019))] 
buoy_location$closest_date[2] <- hurr_Humberto2007$date[which(hurr_Humberto2007$dist_42035 == min(hurr_Humberto2007$dist_42035))] 
buoy_location$closest_date[3] <- hurr_Humberto2007$date[which(hurr_Humberto2007$dist_srst2 == min(hurr_Humberto2007$dist_srst2))] 
buoy_location$closest_date <- lubridate::ymd_hm(buoy_location$closest_date)
buoy_location
```

## Time series

```{r}
hurr_Humberto2007 %<>% mutate(dist_42019 = NA, dist_42035 = NA, dist_srst2 = NA)
for(i in 1:  dim(hurr_Humberto2007)[1]){
  hurr_Humberto2007$dist_42019[i] <- distHaversine(c(hurr_Humberto2007$longitude[i], hurr_Humberto2007$latitude[i]), 
                                                   c(buoy_location$longitude[1], buoy_location$latitude[1]), r=6378137)
  hurr_Humberto2007$dist_42035[i] <- distHaversine(c(hurr_Humberto2007$longitude[i], hurr_Humberto2007$latitude[i]),
                                                   c(buoy_location$longitude[2], buoy_location$latitude[2]), r=6378137)
  hurr_Humberto2007$dist_srst2[i] <- distHaversine(c(hurr_Humberto2007$longitude[i], hurr_Humberto2007$latitude[i]),
                                                   c(buoy_location$longitude[3], buoy_location$latitude[3]), r=6378137)
}
buoy_location %<>% mutate(closest_date = rep(NA, dim(buoy_location)[1]))
# set the closest_date
buoy_location$closest_date[1] <- hurr_Humberto2007$date[which(hurr_Humberto2007$dist_42019 == min(hurr_Humberto2007$dist_42019))] 
buoy_location$closest_date[2] <- hurr_Humberto2007$date[which(hurr_Humberto2007$dist_42035 == min(hurr_Humberto2007$dist_42035))] 
buoy_location$closest_date[3] <- hurr_Humberto2007$date[which(hurr_Humberto2007$dist_srst2 == min(hurr_Humberto2007$dist_srst2))] 
buoy_location$closest_date <- lubridate::ymd_hm(buoy_location$closest_date)

time42019 <- ggplot(subset(buoy_raw, buoy == "buoy_42019"), aes(x = date_time, y = average_wind_speed)) +
  geom_line() +
  geom_rect(aes(xmin=(buoy_location$closest_date[1] - 1*60*60*24),
                xmax=(buoy_location$closest_date[1] + 1*60*60*24),
                ymin=0, ymax=Inf), alpha = .01, fill = "#9898fb") +
  xlab("") + ylab("") + theme_bw()+
  ggtitle("Buoy_42019")

time42035 <- ggplot(subset(buoy_raw, buoy == "buoy_42035"), aes(x = date_time, y = average_wind_speed)) +
  geom_line() +
  geom_rect(aes(xmin=(buoy_location$closest_date[2] - 1*60*60*24),
                xmax=(buoy_location$closest_date[2] + 1*60*60*24),
                ymin=0, ymax=Inf), alpha = .01, fill = "#9898fb") +
  xlab("") + ylab("Average wind speed(hourly)") + theme_bw()+
  ggtitle("Buoy_42035")

timeSRST2<- ggplot(subset(buoy_raw, buoy == "buoy_srst2"), aes(x = date_time, y = average_wind_speed)) +
  geom_line() +
  geom_rect(aes(xmin=(buoy_location$closest_date[3] - 1*60*60*24),
                xmax=(buoy_location$closest_date[3] + 1*60*60*24),
                ymin=0, ymax=Inf), alpha = .01, fill = "#9898fb") +
  xlab("Time") + ylab("") + theme_bw()+
  ggtitle("Buoy_srst2")

grid.arrange(time42019,time42035,timeSRST2,nrow=3)
```

From these figures, we can see that the wind speed at each of the three buoys had a period of rapidly increasing and then decreasing, which indicated that the period is when Hurricane Humberto was passing by each buoy. According to the weather report, the hurricane is notable for its exceptionally rapid intensification near the coast of Texas from a tropical depression into a hurricane within 19 hours. 


### Citation
https://www.weather.gov/hgx/projects_humberto07

https://earthobservatory.nasa.gov/images/19056/hurricane-humberto

https://www.directenergy.com/learning-center/hurricane

https://www.nhc.noaa.gov/data/tcr/AL092007_Humberto.pdf



